Epistemic Semantics in Guarded String Models

### 2020-06-18 tex/draft.tex
Edits at the end of Section 1.

The events in the examples come with pre-conditions. The elevator can
not go up if it is already on floor 3, so $u$ has the pre-condition of
the elevator being of floor 0, 1, or 2.  Similarly $d$ has the
precondition that the elevator is on floor 1, 2 or 3.  Amy can peek at
heads only if the coin is heads up, so $a_1$ has the precondition of
the coin being heads up.  Let $h$ be the Boolean proposition that the
coin is heads up.  In the other example, let $q$ be the proposition
that the elevator is on a high floor (2 or 3), and $p$ be the
proposition that it is on an odd floor (1 or 3). Then preconditions
can be described by Boolean formulas, with $h$ being the precondition
of $a_1$, and $!(pq)$ being the precondition of $u$. Juxtaposition is
used for Boolean conjunction, and the exclamation point for for
Boolean negation. Events come as well with a relation between prior
and following state, for instance with $u$ incrementing the floor.
This is expressed using an operator ``$:$'' (read ``and next'') that
pairs Boolean formulas. The first line in \ex{event1} describes $a_1$
(Amy looking at heads) as happening only in an $h$ state, and as not
changing the state.  Symmetrically, $a_0$ (Amy looking at tails) can
happen only in a not-$h$ state, and does not change the state.  The
third line says that $u$ increments the floor, and can happen only on
floors 0, 1, and 2.  The fourth line describes $d$ in similar terms.
Plus is disjunction.

@@

\begin{example}   \label{event1}
\hspace{-5mm}\( \begin{array}[t]{cl}
a_1 & h:h \\
a_0 & \ev{(!h)}{(!h)} \\
% u & (!q!p) \! : \!(!qp) + \\
u & \ev{(!q!p)}{(!qp)} + \ev{(!qp)}{(q!p)} + \ev{(q!p)}{(qp)} \\
d & \evr{(!q!p)}{(!qp)} + \evr{(!qp)}{(q!p)} + \evr{(q!p)}{(qp)} \\ 
\end{array} \)
\end{example}

@@

Kleene Algebra with Tests is an algebraic theory that is defined by
equations and inequalities, which a has model classes including
guarded string models, relational models, finite models, and matrix
models.  This paper focuses on defining a family of concrete guarded
string algebras, the elements of which are sets of guarded strings.
Definitions and notation mostly follow \cite{kozen2001automata}.
Additional syntax and semantics is included to model multi-agent
epistemic semantics.  Guarded strings over a finite alphabet $E$ are
like ordinary strings, but with truth assignments to a set $T$ of
primitive propositions (primitive tests) alternating with the symbols
from $E$. Following \cite{kozen2001automata}, we call these truth
assignments \emph{atoms}; the set of atoms is notated $\mathcal A$.
In the algebra described by Figure \ref{epikfigure}, $E$ is the set of
events $\{a_1,a_0,b_1,b_0\}$, and in the elevator example, $\{u,d\}$.
In the elevator example, $T$ is $\{p,q\}$, and in the coin example it
is $\{h,t\}$.  We write atoms as the boolean formulae that uniquely
specify the truth assignment. In the elevator example, we have atoms
$\bar q \bar p$, $\bar q p$, $q \bar p$, $q p$, for notational
simplicity we will write these as $\hat0$, $\hat1$, $\hat2$, $\hat3$
respectivelyto indicate the floor each atom represents. In the coin
example, we get $h\bar t$ and $\bar h t$ (for which we use the
shorthand $H$ and $T$), as well as the bogus atoms $\bar h \bar t$ and
$ht$!  To preclude bogosity, the \texttt{constraint} declaration in
Figure~\ref{epikfigure} describes the set of \emph{feasible atoms},
which we parse into a formula $\varphi$ (if there are multiple
\texttt{constraint} declarations, $\varphi$ is the conjunction of the
parsed formulae in each). For a formula $\phi$, the set of feasible atoms is
$\mathcal A_\phi$. In the coin example, the constraint $\varphi$ is
$h\bar t + \bar h t$. Since $\bar h t \models \varphi$ and $h \bar t
\models \varphi$, they are permitted feasible atoms, and since $\bar h
\bar t$ and $h t$ do not satisfy $\phi$ they are not permitted.

@@

Fortunately, Epik provides a syntax to specify the input-output
behavior via the \texttt{event} declarations. From these we construct
a function $P$ from agents to relations on atoms that specify the
effects each action can have. If there are multiple event declarations
for the same event, only the latter one is incorporated. To construct
the effect relation $P(e)$ for event $e$, we add a pair $\alpha,
\beta$ to the relation $P(e)$ for every disjunct $\phi:\phi'$ such
that $\alpha \models \phi$ and $\beta \models \phi'$.  These have the
status of event types, in that they can ``happen'' in different
possible worlds. Here happening corresponds to incrementing the world
with some compatible element of the event type.  Where $e$ is a bare
event symbol, $\llbracket e \rrbracket^{\varphi,P}$ is the
corresponding set of guarded strings, consisting of the bare event
decorated with compatible truth assignments denoted by $P$ that are
admitted by the world-constraints $\varphi$.  See this formally in
Figure~\ref{fig:interp}, and the examples in \ex{events1}.
% u (!q!p):(!qp) + (!qp):(q!p) + (q!p):(qp) 173
% d (!qp):(!q!p) + (q!p):(!qp) + (qp):(q!p)

>>>>

The events in the examples come with pre-conditions. The elevator can
not go up if it is already on floor 3, so $u$ has the pre-condition of
the elevator being of floor 0, 1, or 2.  Similarly $d$ has the
precondition that the elevator is on floor 1, 2 or 3.  Amy can peek at
heads only if the coin is heads up, so $a_1$ has the precondition of
the coin being heads up.  Let $h$ be the Boolean proposition that the
coin is heads up.  In the other example, let $q$ be the proposition
that the elevator is on a high floor (2 or 3), and $p$ be the
proposition that it is on an odd floor (1 or 3). Then preconditions
can be described by Boolean formulas, with $h$ being the precondition
of $a_1$, and $\overline{pq}$ being the precondition of $u$. Juxtaposition is
used for Boolean conjunction, and the overbar for
Boolean negation. Events come as well with a relation between prior
and following state, for instance with $u$ incrementing the floor.
This is expressed using an operator ``$:$'' (read ``and next'') that
pairs Boolean formulas. The first line in \ex{event1} describes $a_1$
(Amy looking at heads) as happening only in an $h$ state, and as not
changing the state.  Symmetrically, $a_0$ (Amy looking at tails) can
happen only in a not-$h$ state, and does not change the state.  The
third line says that $u$ increments the floor, and can happen only on
floors 0, 1, and 2.  The fourth line describes $d$ in similar terms.
Plus is disjunction.

@@

\begin{example}   \label{event1}
\hspace{-5mm}\( \begin{array}[t]{cl}
a_1 & h:h \\
a_0 & \ev{\bar{h}}{\bar{h}} \\
% u & (\bar{q}\bar{p}) \! : \!(\bar{q}p) + \\
u & \ev{(\bar{q}\bar{p})}{(\bar{q}p)} + \ev{(\bar{q}p)}{(q\bar{p})} + \ev{(q\bar{p})}{(qp)} \\
d & \evr{(\bar{q}\bar{p})}{(\bar{q}p)} + \evr{(\bar{q}p)}{(q\bar{p})} + \evr{(q\bar{p})}{(qp)} \\ 
\end{array} \)
\end{example}

In this discussion, a sequence such as $\bar{q}{p}$ can be viewed as a
valuation of the primitive propositions $q$ and $p$, which is used to
describe world state. The primitives are listed in fixed order, and
left unmarked (indicating true) or marked with the overbar (indicating
false).  Suppose that in the coin example, we had an additional
primitive stative proposition (or atomic test) $t$, interpreted as
tails.  Since a coin is heads or tails but not both, we want to allow
the valuations $h\bar{t}$ and $\bar{h}t$, and disallow $ht$ and
$\bar{h}\bar{t}$.  This is enforced by a {\em state formula},
which is a simply Boolean formula, in this case \exi{effect}{a}.
Where
{\sf{B}} is a set of atomic tests and $\rho$ is a state constraint
over {\sf{B}}, ${\cal{A}}_{\mbox{\sf\tiny{B}}}^{\rho}$ is the set of valuations of
{\sf{B}} that make formula $\rho$ true.

Formulas like the ones in \ex{event1} that describe pre- and
post-conditions are {\em effect formulas}. They are interpreted as
defining relations between valuations, as defined in Figure
\ref{effectfigure}.  The valuations they relate are constrained by the
state formula as well.  For the heads-tails example, let the state
formula and the effect formula for $a_1$ (Aly peeking at heads) be as
specified in \ex{effect}. Then the set of valuations determined by
$\rho$ and the relation on valuations for the event $u$ are \sem{rho}
and $\sem{\zeta}^{\sem{rho}}$ as given at the bottom in \ex{6}.  Given
a state formula $\rho$, the the relation characterizing pre- and
post-conditions of an event with effect formula $\zeta$ is defined to
be $\sem{\zeta}^{\sem{\rho}}$.

\begin{example}   \label{effect}
\begin{tabular}[t]{l}
State formula ($\rho$)  $h\bar{t} + \bar{h}t$ \\
Effect formula for $a_1$ ($\zeta$)  $\ev{h}{h}$ \\
\( \sem{\rho} = \{h\bar{t},\bar{h}t\} \)  \\
\( \sem{\zeta}^{\sem{\rho}} = \{\pair{h\bar{t}}{h\bar{t}}\} \)  \\
\end{tabular}
\end{example}

\begin{figure}
\hspace{-2mm}\begin{tabular}{lll}
  state formulas & $\rho,\sigma...$ & $\rho{:}{:}{=} \, \begin{array}[t]{l}\tuple{\mbox{atomic tests}}\,|\\
    0\,|\,1\,| \; \rho{+}\sigma\;| \\ \rho\,{\&}\,\sigma\;|\;\bar{\rho}\end{array}$ \\
  effect formulas & $\zeta,\eta...$ & $\zeta{:}{:}{=} \, \begin{array}[t]{l} \rho\,{:}\,\sigma \,|\, \zeta{+}\eta \\
    \zeta\,{\&}\,\eta\;|\;\bar{\zeta}\end{array}$
\end{tabular}
\vspace{1mm}

\(
\begin{array}{cll}
  \sem{\rho} & = & \mbox{Set of Boolean valuations of} \\
            && \mbox{the atomic tests that verify $\rho$.} \\
\sem{\rho\,{:}\,\sigma}^V & = & (\sem{\rho}\cap V)\times(\sem{\sigma}\cap V)\\
\sem{\zeta{+}\eta}^V & = & \sem{\zeta}^V\cup\sem{\eta}^V\\
\sem{\zeta{\&}\eta}^V & = & \sem{\zeta}^V\cap\sem{\eta}^V\\
\sem{\bar{\zeta}}^V & = & V \times V \setminus \sem{\zeta}^V\\
\end{array}  
\)

\caption{Syntax and semantics of state formulas and effect formulas. Effect formulas
  are evaluated relative to as set Boolean valuations $V$, and denote relations between
  Boolean valuations.}
\label{effectfigure}
\end{figure}

@@

Kleene Algebra with Tests is an algebraic theory that is defined by
equations and inequalities, which a has model classes including
guarded string models, relational models, finite models, and matrix
models.  This paper focuses on defining a family of concrete guarded
string algebras, the elements of which are sets of guarded strings.
Definitions and notation mostly follow \cite{kozen2001automata}.
Additional syntax and semantics is included to model multi-agent
epistemic semantics.  Guarded strings over a finite alphabet {\sf{P}} are
like ordinary strings, but with truth assignments to a set $T$ of
primitive propositions (primitive tests) alternating with the symbols
from $E$. Following \cite{kozen2001automata}, we call these truth
assignments \emph{atoms}; the set of atoms is notated $\mathcal A$.
In the algebra described by Figure \ref{epikfigure}, $E$ is the set of
events $\{a_1,a_0,b_1,b_0\}$, and in the elevator example, $\{u,d\}$.
In the elevator example, $T$ is $\{p,q\}$, and in the coin example it
is $\{h,t\}$.  We write atoms as the boolean formulae that uniquely
specify the truth assignment. In the elevator example, we have atoms
$\bar q \bar p$, $\bar q p$, $q \bar p$, $q p$, for notational
simplicity we will write these as $\hat0$, $\hat1$, $\hat2$, $\hat3$
respectivelyto indicate the floor each atom represents. In the coin
example, we get $h\bar t$ and $\bar h t$ (for which we use the
shorthand $H$ and $T$), as well as the bogus atoms $\bar h \bar t$ and
$ht$!  To preclude bogosity, the \texttt{constraint} declaration in
Figure~\ref{epikfigure} describes the set of \emph{feasible atoms},
which we parse into a formula $\varphi$ (if there are multiple
\texttt{constraint} declarations, $\varphi$ is the conjunction of the
parsed formulae in each). For a formula $\phi$, the set of feasible atoms is
$\mathcal A_\phi$. In the coin example, the constraint $\varphi$ is
$h\bar t + \bar h t$. Since $\bar h t \models \varphi$ and $h \bar t
\models \varphi$, they are permitted feasible atoms, and since $\bar h
\bar t$ and $h t$ do not satisfy $\phi$ they are not permitted.

@@

Fortunately, Epik provides a syntax to specify the input-output
behavior via the \texttt{event} declarations. From these we construct
a function $P$ from agents to relations on atoms that specify the
effects each action can have. If there are multiple event declarations
for the same event, only the latter one is incorporated. To construct
the effect relation $P(e)$ for event $e$, we add a pair $\alpha,
\beta$ to the relation $P(e)$ for every disjunct $\phi:\phi'$ such
that $\alpha \models \phi$ and $\beta \models \phi'$.  These have the
status of event types, in that they can ``happen'' in different
possible worlds. Here happening corresponds to incrementing the world
with some compatible element of the event type.  Where $e$ is a bare
event symbol, $\llbracket e \rrbracket^{\varphi,P}$ is the
corresponding set of guarded strings, consisting of the bare event
decorated with compatible truth assignments denoted by $P$ that are
admitted by the world-constraints $\varphi$.  See this formally in
Figure~\ref{fig:interp}, and the examples in \ex{events1}.
% u (\bar{q}\bar{p}):(\bar{q}p) + (\bar{q}p):(q\bar{p}) + (q\bar{p}):(qp) 173
% d (\bar{q}p):(\bar{q}\bar{p}) + (q\bar{p}):(\bar{q}p) + (qp):(q\bar{p})
